Ages of Stars and Planets Observed by Kepler

### 2024-04-23 paper/ms.tex
update ms

Owing to the success of the transit and radial velocity techniques,
exoplanet science is going through an age of abundance in which we are
beginning to explore thousands of new worlds that orbit in close
proximity to their host stars.
However, most of these exoplanets are old, with ages comparable to
that of the solar system.
This limits our ability to understand their origins, and in particular
to study processes such as dynamical migration (CITE), thermal cooling
(CITE), and atmospheric loss (CITE) that typically happen soon after
disk dispersal.
Finding and characterizing young ($<$1\,Gyr) exoplanets is one way to
advance our understanding of how and when these processes occur.

@@

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods have been reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
and therefore measurements of the \ion{Li}{1} 6708\,\AA\ doublet, in
general have only been acquired for the Kepler objects of interest
(KOIs), which comprise a few percent of Kepler's targets.
In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).

@@

We measured the lithium equivalent widths using a procedure
adapted from previous work \citep{Bouma_2021}.  Our stars of interest
are FGK stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $\gamma$$\approx$9\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA (we considered 10\,\AA\ and 20\,\AA; the results were
broadly consistent).  We continuum-normalized by fitting a
third-degree Chebyshev series, while excluding regions with
absorption lines.  We then numerically integrated the resulting
spectrum using a one-component Gaussian with free amplitude, width,
and mean, and estimated uncertainties on the line width through a
Monte Carlo procedure that bootstraped against the local scatter in
the continuum.

@@

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the issue
that, especially over the first gigayear of stellar spin-down, stars
with the same mass and same rotation period can have very different
ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
posterior should therefore incorporate this intrinsic population-level
scatter into the precision with which the age can be inferred.
Figure~\ref{fig:models} exemplifies this problem, particularly in
regions where the 0.1-1\,Gyr stars overlap.

@@

	\caption{
		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
		Planets highlighted in this plot have $t$<1\,Gyr at $2$$\sigma$.
%		
%		Two possible selection functions for what defines a young, sub-gigayear, planet
%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
%		Circles are previously reported measurements from the literature.
		Stars are newly calculated ages for Kepler planets from this work; circles
		are planets from the literature meeting the same requirement.
		Gray points are transiting planets from the NASA Exoplanet Archive older than
		1\,Gyr.
		The three age bins span 10-46\,Myr, 46-215\,Myr, and 215-1000\,Myr.
		\label{fig:rp_period_age_results}
	}
\end{figure*}

>>>>

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars (CITE, CITE,
CITE).
However, most known exoplanets have ages comparable to that of the solar
system.
This unfortunate fact leaves many gaps in our knowledge of
exoplanetary origins, particularly regarding to degree to which
dynamical migration (CITE), thermal cooling (CITE), and atmospheric
loss (CITE) sculpt planetary systems after the disk disperses.
Finding and characterizing young ($<$1\,Gyr) exoplanets offers one
possible workaround, since these objects can show both individual (CITE,
CITE) and population-level (CITE) evidence for these types of
evolutionary processes.

@@

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods have been reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
which can yield measurements of the \ion{Li}{1} 6708\,\AA\ doublet, 
have only been acquired for the Kepler objects of interest
(KOIs), which comprise a few percent of Kepler's targets.
In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).

@@

We measured the lithium equivalent widths using a procedure
adapted from previous work \citep{Bouma_2021}.  Our stars of interest
are FGK stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $\gamma$$\approx$9\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we considered 10\,\AA\ and 20\,\AA; the results were
broadly consistent).  We continuum-normalized by fitting a
third-degree Chebyshev series, while excluding regions with
absorption lines.  We then numerically integrated the resulting
spectrum using a one-component Gaussian with free amplitude, width,
and mean, and estimated uncertainties on the line width through a
Monte Carlo procedure that bootstraped against the local scatter in
the continuum.

@@

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the issue
that, especially over the first gigayear of stellar spin-down, stars
with the same mass and same rotation period can have very different
ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
posterior should therefore incorporate this intrinsic population-level
scatter into the precision with which the age can be inferred.
Figure~\ref{fig:models} highlights this problem, particularly in
regions where the 0.1-1\,Gyr stars overlap.

@@

	\caption{
		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
		Planets highlighted in this plot have ages $t$<1\,Gyr at $2$$\sigma$.
%		
%		Two possible selection functions for what defines a young, sub-gigayear, planet
%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
%		Circles are previously reported measurements from the literature.
		Stars denote systems that are newly characterized in this work; circles
		are planets from the literature meeting the same requirement.
		Gray points are transiting planets from the NASA Exoplanet Archive older than
		1\,Gyr.
		The selection function matters:
		if we were to require $t$<1\,Gyr and $t/\sigma_{\rm t}>3$, we would loose $\approx$ten
		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
		\label{fig:rp_period_age_results}
	}
\end{figure*}
